Problem of Step 4 (debias) of fGFPCA on NHANES data

### 2024-02-13 Docs/NhanesApplExample.Rmd
reversed eigenvalue problem: simulation

# Fit fGFPCA model

```{r data_subset}
# head(df)
df <- df %>% rename(id=SEQN, Y=Z)
N <- length(unique(df$id))
J <- length(unique(df$sind))
```

# Step 1: Bin data

- Bin every 10 observations

```{r, message=FALSE}
# bin data
bin_w <- 10 # bin width
n_bin <- J/bin_w # number of bins
brks <- seq(0, J, by = bin_w) # cutoff points
mid <- (brks+bin_w/2)[1:n_bin] # mid points

df$bin <- cut(df$sind, breaks = brks, include.lowest = T, labels = mid)
df$bin <- as.numeric(as.character(df$bin))
```

# Step 2-4:  Local GLMM, fPCA, splines interpolation and GLMM reevaluation

I'd like to investigate with different sample sizes.

```{r, cache=TRUE, results='hide'}
# sample size 
N_sub <- rep(500, 4)
fpca_model_list <- list()
interp_pc_est <- list()
debias_model_list <- list()
time_vec <- rep(NA, length(N_sub))

pb <- txtProgressBar(min=0, max = length(N_sub), style = 3)
for(i in seq_along(N_sub)){

  # training sample of a certain size
  Ntr <- N_sub[i]
  train_id <- sample(unique(df$id), size = Ntr) 
  train_df <- df %>% filter(id %in% train_id)

  t1=Sys.time()
  # local GLMM
  train_bin_lst <- split(train_df, f = train_df$bin)
  df_est_latent <- lapply(train_bin_lst, function(x){pred_latent(x, n_node = 0)}) 
  df_est_latent <- bind_rows(df_est_latent)

  # FPCA
  uni_eta_hat <- df_est_latent %>% filter(bin==sind)
  mat_est_unique <- matrix(uni_eta_hat$eta_hat, nrow=Ntr, ncol=n_bin, byrow = F) 
  fpca_mod <- fpca.face(mat_est_unique, argvals = mid, var=T, npc=K)
  fpca_model_list[[i]] <- fpca_mod

  # interpolation
  p <- 3 # order of b splines 
  knots <- 35 # number of knots (same from FPCA model)
  knots_values <- seq(-p, knots + p, length = knots + 1 + 2 *p)/knots
  knots_values <- knots_values * (max(mid) - min(mid)) + min(mid)

@@

  df_phi <- matrix(NA, J, K) 
  for(k in 1:K){
    lm_mod <- lm(fpca_mod$efunctions[,k] ~ B-1)
    df_phi[,k] <- Bnew %*% coef(lm_mod)
  }# project binned eigenfunctions onto the original grid
  interp_pc_est[[i]] <- df_phi

  # debias model
  df_phi <- data.frame(sind = 1:J, df_phi)
  colnames(df_phi) <- c("sind", paste0("phi", 1:K))
  train_df <- train_df %>% 
    left_join(df_phi, by = "sind")
  train_df$id <- as.factor(train_df$id)
  debias_glmm <- bam(Y ~ s(sind, bs="cr")+
                     s(id, by=phi1, bs="re")+
                     s(id, by=phi2, bs="re")+
                     s(id, by=phi3, bs="re")+
                     s(id, by=phi4, bs="re"), 
                   family = binomial, 
                   data=train_df, 
                   method = "fREML",
                   discrete = TRUE)
  debias_model_list[[i]] <- debias_glmm
  t2 <- Sys.time()

  time_vec[i] <- as.numeric(t2-t1, unit = "secs")

  setTxtProgressBar(pb, i)
}

close(pb)
```

<!-- - Time spent on model fitting -->

```{r, eval=FALSE}
time_vec
```

<!-- - FPCA eigenvalues -->

```{r, eval=FALSE}
lapply(fpca_model_list, function(x){x$evalues})
lapply(fpca_model_list, function(x){x$pve})
```

- Eigenvalues from debias models

```{r}
lapply(debias_model_list, function(x){1/x$sp[2:5]})
```

```{r}
pc_plot_list <- list()

for(i in  1:length(N_sub)){

  df_phi1 <- data.frame(t=mid, fpca_model_list[[i]]$efunctions)
  df_phi2 <- data.frame(t = 1:J, interp_pc_est[[i]])

  this_plot <- left_join(df_phi2, df_phi1, by = "t") %>%
    ggplot()+
    geom_point(aes(x=t, y=X1.y, col = "PC1"), na.rm = T, size = 0.5)+
    geom_point(aes(x=t, y=X2.y, col = "PC2"), na.rm = T, size = 0.5)+
    geom_point(aes(x=t, y=X3.y, col = "PC3"), na.rm = T, size = 0.5)+
    geom_point(aes(x=t, y=X4.y, col = "PC4"), na.rm = T, size = 0.5)+
    geom_line(aes(x=t, y=X1.x, col = "PC1"))+
    geom_line(aes(x=t, y=X2.x, col = "PC2"))+
    geom_line(aes(x=t, y=X3.x, col = "PC3"))+
    geom_line(aes(x=t, y=X4.x, col = "PC4"))+
    labs(x="time", y = "", title = paste0("Iter", i))

  pc_plot_list[[i]] <- this_plot

}

ggarrange(plotlist = pc_plot_list, common.legend = T)
```

The "reverse eigenvalue problem" does not always happen. When training sample size is small (like maybe 200), the problem happened less. So this is definately a numeric problem. But a very small sample also would induce sigularity issues.

Some other thoughts:

1. What about estimation method of smooth parameter? It seems GCV doesn't treat model as mixed effects. REML may be too slow without the "discrete=T" option on the full dataset. 
2. Is is possible to impose monotone constraint on smoothing parameter estimation?

>>>>

```{r bin_data}
# head(df)
df <- df %>% rename(id=SEQN, Y=Z)
N <- length(unique(df$id))
Ntr <- 500
J <- length(unique(df$sind))

load(here("Data/Appl_debias_model.RData"))

# parameters used in binning
bin_w <- 10 # bin width
n_bin <- J/bin_w # number of bins
brks <- seq(0, J, by = bin_w) # cutoff points
mid <- (brks+bin_w/2)[1:n_bin] # mid points
```

# Estimates

- Eigenvalues

```{r evals}
evals <- 1/debias_glmm$sp[2:5]/n_bin
```

```{r}
df_phi <- debias_glmm$model %>% select(sind, starts_with("phi")) %>% distinct(.) %>%
  mutate_at(vars(starts_with("phi")), function(x){x*sqrt(n_bin)})

df_phi %>% pivot_longer(2:5) %>%
  ggplot(aes(x=sind, y=value))+
  geom_line()+
  facet_wrap(~name)
```

- Mean function

```{r}
mu <- predict(debias_glmm, type = "terms")[1:J, 1]
plot(1:J, mu, type = "l")
```

# Simulation using re-evaluated estimates

```{r gen_data}
# generate score
xi <- rmvnorm(Ntr, mean = rep(0, K), sigma = diag(evals))
xi %>% data.frame() %>% pivot_longer(1:4, names_to = "PC", values_to = "score") %>%
  ggplot()+
  geom_boxplot(aes(x=PC, y=score))+
  geom_jitter(aes(x=PC, y=score), size = 0.5)

# individual random effects
bi <- as.matrix(df_phi[, 2:5]) %*% t(xi)

# add in mean 
eta <- apply(bi, 2, function(x){x+mu})
gen_df <- data.frame(id = factor(rep(1:Ntr, each=J)),
                     sind = rep(1:J, Ntr), 
                     eta_i = as.vector(eta))

# binary outcome
gen_df$Y <- rbinom(Ntr*J, size=1, prob=plogis(gen_df$eta_i))

gen_df %>% filter(id %in% 1:4) %>%
  ggplot()+
  geom_point(aes(x=sind, y=Y), size = 0.5)+
  geom_line(aes(x=sind, y=plogis(eta_i)), col = "red")+
  facet_wrap(~id)+
  labs(title = "Generated data")
```

# fGFPCA on generated data

```{r fGFPCA1}
# bin 
gen_df$bin <- cut(gen_df$sind, breaks = brks, include.lowest = T, labels = mid)
gen_df$bin <- as.numeric(as.character(gen_df$bin))

# local GLMM
train_bin_lst <- split(gen_df, f = gen_df$bin)
df_est_latent <- lapply(train_bin_lst, function(x){pred_latent(x, n_node = 0)}) 
df_est_latent <- bind_rows(df_est_latent)

df_est_latent %>% filter(id %in% 1:4) %>%
  ggplot()+
  geom_point(aes(x=sind, y=Y), size = 0.5)+
  geom_line(aes(x=sind, y=plogis(eta_i), col = "True"))+
  geom_line(aes(x=sind, y=plogis(eta_hat), col = "Estimate"))+
  facet_wrap(~id)+
  labs(title = "Generated data")
```

```{r fGFPCA2}
# FPCA
uni_eta_hat <- df_est_latent %>% filter(bin==sind)
mat_est_unique <- matrix(uni_eta_hat$eta_hat, nrow=Ntr, ncol=n_bin, byrow = F) 
fpca_mod <- fpca.face(mat_est_unique, argvals = mid, var=T, npc=K)
```

```{r re_eval_PC}
# interpolation
p <- 3 # order of b splines 
knots <- 35 # number of knots (same from FPCA model)
knots_values <- seq(-p, knots + p, length = knots + 1 + 2 *p)/knots
knots_values <- knots_values * (max(mid) - min(mid)) + min(mid)

@@

est_phi <- matrix(NA, J, K) 
for(k in 1:K){
  lm_mod <- lm(fpca_mod$efunctions[,k] ~ B-1)
  est_phi[,k] <- Bnew %*% coef(lm_mod)
} 
colnames(est_phi) <- paste0("phi", 1:K)
```

I would like to look at estimated PC functions:

```{r}
# PC functions after re-evaluation
bind_rows(
  data.frame(sind=1:J, est_phi*sqrt(n_bin), type = "re-evaluated"),
  data.frame(df_phi, type = "True")
) %>%
  pivot_longer(2:5) %>%
  ggplot()+
  geom_line(aes(x=sind, y=value, col=type))+
  facet_wrap(~name)
```

It looks like something went wrong in the estimation of PC3. The estimated line is either identical or off by a sign compared to the true PC functions. That said, PC2 seems a bit off too.

What about the estimated PC functions from FPCA without re-evaluation?

```{r}
fpca_phi <- fpca_mod$efunctions*sqrt(n_bin)
colnames(fpca_phi) <- paste0("phi", 1:K)
bind_rows(
  data.frame(df_phi, type = "True"),
  data.frame(sind = mid, fpca_phi, type = "FPCA")
) %>%
  pivot_longer(2:5) %>%
  ggplot()+
  geom_point(aes(x=sind, y=value, col=type),size = 0.5)+
  facet_wrap(~name)
```

Looks like FPCA did not capture the shape of PC2 and PC3.

Next I'd like to re-evaluate the eigenvalues

```{r}
# debias model
gen_df <- gen_df %>% 
  left_join(df_phi, by = "sind")
gen_df$id <- as.factor(gen_df$id)
debias_glmm2 <- bam(Y ~ s(sind, bs="cr")+
                   s(id, by=phi1, bs="re")+
                   s(id, by=phi2, bs="re")+
                   s(id, by=phi3, bs="re")+
                   s(id, by=phi4, bs="re"),
                 family = binomial,
                 data=gen_df,
                 method = "fREML",
                 discrete = TRUE)
```

Eigenvalues before and after re-evaluation

```{r, class.source='fold-show'}
# fpca
fpca_mod$evalues/n_bin

# re-evaluated
1/debias_glmm2$sp[2:5]
```

It looks like the flip happened again.

At this point, I think we could say the reversed eigenvalue problem is an artifact of the dataset it self. For some reason, FPCA was not able to estimated PC2 and PC3 very well, and the GLMM model later on impose more penalty on the second random term.
